Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis

### 2018-03-25 rigid_bodies.tex
Small changes

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter. As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$. Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$. It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$. Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates. Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$. In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$. In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$. Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$.

@@

Speak about the Generalized Equipartition Theorem....

>>>>

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter. As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

@@

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$. It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$. Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated. If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates. Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$. In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$. In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$. Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.

@@

Tolman's generalized equipartition theorem \cite{Uline_2008}
\begin{equation*}
\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T
\end{equation*}
